---
title: "Two-Way ANOVA"
highlight-style: solarized
code-block-border-left: "#FCC251"

format: 
  html: default
  gfm: default
---

```{css, echo = FALSE, purl = FALSE}
h1, h2, h3, h4 {
  color: #CB6015;
}
```

```{=html}
<style> 
.math.display .MathJax {
  font-size: 95% !important;
}
</style>
```

```{r echo = FALSE, eval = FALSE}


## Thompson, M., Lie, Y. & Green, S. (2023). Flexible structural equation modeling
## approaches for analyzing means. In R. Hoyle (Ed.), Handbook of structural
## equation modeling (2nd ed., pp. 385-408). New York, NY: Guilford Press.
```

Thompson, M., Lie, Y. & Green, S. (2023). Flexible structural equation modeling approaches for analyzing means. In R. Hoyle (Ed.), *Handbook of structural equation modeling* (2nd ed., pp. 385-408). New York, NY: Guilford Press.

<br />

This example shows the SEM approach to Part 3: Two-way ANOVA. Results are reported in Tables 21.3 and 21.4 (pp. 395, 396).

The data file (`satisfactionI.csv` in the `data` folder) needs manipulation before it can be used: the format needs to be changed from "long" to "wide"; and the Gender X Coping Strategy interaction needs a grouping variable set up. These manipulations are completed in `ANOVA_data.r`.

<br />

#### Load relevant packages and get the data

Load the relevant packages, and run `ANOVA_data.r` to get the data.

```{r echo = c(-7, -2:-1), eval = FALSE}

## Load packages
library(lavaan)
library(here)         # Relative paths
library(DescTools)    # Cramer's V

## Get the data
path <- here::here("Green_2023", "data", "ANOVA_data.r")
source(path)
head(df)
```

The variables used in this example are:

 - x - Coping Strategy ("a" - no strategy; "b" - discussion; "c" - exercise) 
 - g - Gender
 - y - dependent variable ("after" self-satisfaction scores)
 - sg - Gender X Coping Strategy interaction

<br />

#### Preliminary results - Cramer's V

On page 394, TLG give Cramer's V for the Gender X Coping Strategy crosstabulation. As far as I know, Cramer's V is not available in base R, but **DescTools** is one of possibly many packages that has a function for Cramer's V.

```{r echo = -3:-1, eval = FALSE}

## Cramer's V
## Check with page 394
DescTools::CramerV(df$g, df$x)
```

However, it is easy to calculate Cramer's V without the need for the extra package, given the formula:

$$
 \mathsf{Cramer's ~ V} = \sqrt{\frac{\upchi^2 / n}{\min(r-1, ~ c-1)}}
$$

where $n$ is the sample size, $r$ is the number of rows, and $c$ is the number of columns.

```{r echo = -2:-1, eval = FALSE}

## Cramer's V by hand
chisq <- unname(chisq.test(df$g, df$x)$statistic)
n <- length(df$g)           # Sample size
r <- length(unique(df$g))   # Number of rows
c <- length(unique(df$x))   # Number of columns

CV <- sqrt((chisq/n)/min(r-1, c-1)); CV
```

Standardised residuals will give the direction of the relationship (p. 394).

```{r echo = -2:-1, eval = FALSE}

## Direction of the relationship
chisq.test(df$g, df$x)$stdres
```

<br />

#### Preliminary results - Gender X Coping Strategy crosstabulation

Table 21.3 (p. 395) gives the cell means and frequecies, and the weighted and unweighted marginal means.

Get the cell means and frequencies.

```{r echo = -3:-1, eval = FALSE}

## Cell means and cell frequencies
## Check cell means and frequencies in Table 21.3
means <- tapply(df$y, list(df$g, df$x), mean); means     # Cell means
freq <- table(df$g, df$x); freq                          # Cell frequencies
```

Get the unweighted and weighted marginal means.

```{r echo = -2:-1, eval = FALSE}

## Check unweighted and weighted means in Table 21.3
# Unweighted marginal means
apply(means, 1, mean)      # Gender
apply(means, 2, mean)      # Coping Strategy
 
# Weighted marginal means
tapply(df$y, df$g, mean)     # Gender
tapply(df$y, df$x, mean)     # Coping Strategy
```

<br />
   
### Structural Equation Modeling  using **lavaan**

The SEM model for two-way ANOVA is shown below. The diagram shows the "Less Constrained" model - the six means, represented by the label on the arrows connecting the "1" to the dependent variable, differ. To be consistent with the ANOVA assumption of homogeneity of variances, the residual variances are constrained to be equal.

![](images/two_way_ANOVA.svg){fig-align="left"}

The model statements are shown below. The "Less Constrained" model allows the means (represented by the labels, am, af, ..., cf) to differ across the groups. The constraints statements are added to the "Less Constrained" statement to give the "More Constrained" models. The "More Constrained" models are contrasted with the "Less Constrained" model to test for the Gender and Coping Strategy main effects (weighted and unweighted) and the Gender X Coping Strategy interaction. In each case the residual variances are constrained to equality.

Constraint for the unweighted Gender main effect - Restrict the mean for males to equal the mean for females. But there are three means for females, one for each Coping Strategy group. Similarly, there are three means for males. Simply constrain the sum of the three means for males to equal the sum of the three means for females.

Constraints for the unweighted Coping Strategy main effect - Restrict the mean for “a” strategy to equal the mean for “b” strategy to equal the mean for “c” strategy. That is, constrain the sum of the two “a” means to equal the sum of the two “b” means; and the sum of the two “b” means to equal the sum of the two “c” means.

To test for the main effects applied to weighted means, the constraints are set the same way as before except the means are weighted in proportion to the cell frequencies.

Constraints for the Gender X Coping Strategy interaction - The “More Constrained” model needs the means to be constrained so that the difference between the mean for “female” and the mean for “male” remains constant across levels of “Coping Strategy”. That is: the difference between “female” mean and “male” mean for the “a” strategy equals the difference between “female” mean and “male” mean for the “b” strategy; and the difference between “female” mean and “male” mean for the “b” strategy equals the difference between “female” mean and “male” mean for the “c” strategy.


```{r echo = c(-2:-1), eval = FALSE}

### Two-way ANOVA - SEM
# Less Constrained model
lc <- "y ~ c(am, af, bm, bf, cm, cf)*1       # Means
       y ~~ c(e, e, e, e, e, e)*y            # Variances"

lc.fit <- sem(lc, data = df, group = "sg")
summary(lc.fit)


# Gender main effect - unweighted means
constraints <- "af + bf + cf == am + bm + cm"
gend_unw <- c(lc, constraints)

gend_unw.fit <- sem(gend_unw, data = df, group = "sg")
summary(gend_unw.fit)

anova(gend_unw.fit, lc.fit)   # Compare the two models


# Coping Strategy main effect - unweighted means
constraints <- "
   af + am == bf + bm 
   af + am == cf + cm"
strat_unw <- c(lc, constraints)

strat_unw.fit <- sem(strat_unw, data = df, group = "sg")
summary(strat_unw.fit)

anova(strat_unw.fit, lc.fit)   # Compare the two models


# Gender main effect - weighted means
freq                     # To assist with constructing constraints
constraints <- "(3*af + 3*bf + 6*cf)/12 == (6*am + 3*bm + 3*cm)/12"
gend_w = c(lc, constraints)

gend_w.fit <- sem(gend_w, data = df, group = "sg")
summary(gend_w.fit)

anova(gend_w.fit, lc.fit)   # Compare the two models


# Coping Strategy main effect - weighted means
# Compare with SEM section in Table 21.4
freq
constraints <- "
   (3*af + 6*am)/9 == (3*bf + 3*bm)/6 
   (3*bf + 3*bm)/6 == (6*cf + 3*cm)/9"
strat_w <- c(lc, constraints)

strat_w.fit <- sem(strat_w, data = df, group = "sg")
summary(strat_w.fit)

anova(strat_w.fit, lc.fit)   # Compare the two models


# Gender X Coping Strategy interaction
constraints <- "
   (af - am) == (bf - bm) 
   (bf - bm) == (cf - cm)"
inter <- c(lc, constraints)

inter.fit <- sem(inter, data = df, group = "sg")
summary(inter.fit)

anova(inter.fit, lc.fit)     # Compare the two models
```

<br />

```{r, echo = FALSE, include = FALSE, eval = TRUE, purl = FALSE}
input <- knitr::current_input()
output <- "03_two_way_ANOVA.r"
knitr::purl(input, output, documentation = 0, quiet = TRUE)
```

The R script with minimal commenting is available in [`{r} noquote(output)`](`{r} noquote(output)`).
